---
title: "eTable 1. Category-level deaths and per-capita rates, 2021 and 2024"
format:
  html:
    embed-resources: true
    page-layout: full
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(here)
library(gt)
library(arrow)
source(here::here("code", "utils.R"))

cats_self <- c(
    "speeding_only", "alcohol_only", "unrestrained_only", "speed_alc",
    "speed_unr", "alc_unr", "all_three", "none_flagged"
)
cats_other <- c("pedestrian", "bicyclist", "other_nonpv")
cats_all <- c(cats_self, cats_other)

## Observed national death counts + population (deterministic). The rates
## carry the uncertainty, the counts are the data.
nat <- arrow::read_parquet(here::here("data", "panel_state_year.parquet")) |>
    dplyr::filter(year %in% c(2021, 2024)) |>
    dplyr::group_by(year) |>
    dplyr::summarise(population = sum(population),
        dplyr::across(dplyr::all_of(cats_all), sum), .groups = "drop") |>
    dplyr::mutate(
        behavioral = rowSums(
            dplyr::across(dplyr::all_of(return_behavioral_flagged()))
        ),
        other = rowSums(dplyr::across(dplyr::all_of(cats_other))),
        total_crude = rowSums(dplyr::across(dplyr::all_of(cats_all)))
    )

pop_2021 <- nat$population[nat$year == 2021]
pop_2024 <- nat$population[nat$year == 2024]

deaths_n <- function(comp, yr) {
    nat[[comp]][nat$year == yr]
}

## Rate uncertainty intervals from the negative-binomial Monte Carlo.
cr <- arrow::read_parquet(
    here::here("data", "summaries", "category_rates_ci.parquet")
) |>
    dplyr::filter(geography == "national")
n_iter <- max(cr$n)
cr <- cr |> dplyr::select(component, metric, mean, p025, p975)

get_ci <- function(comp, met) {
    row <- cr |> dplyr::filter(component == comp, metric == met)
    list(mean = row$mean, lower = row$p025, upper = row$p975)
}
```

```{r table-data}
## Non-breaking spaces so gt preserves indentation in HTML (same as Table 1).
INDENT2 <- strrep("\u00a0", 4)
INDENT4 <- strrep("\u00a0", 8)

rate_row <- function(label, comp, indent = "") {
    a <- get_ci(comp, "rate_a")
    b <- get_ci(comp, "rate_b")
    d <- get_ci(comp, "rate_change")
    tibble::tibble(
        Category = paste0(indent, label),
        d2021 = fmt_int(deaths_n(comp, 2021)),
        r2021 = fmt_rate_ci(a$mean, a$lower, a$upper),
        d2024 = fmt_int(deaths_n(comp, 2024)),
        r2024 = fmt_rate_ci(b$mean, b$lower, b$upper),
        change = fmt_val_ci(d$mean, d$lower, d$upper)
    )
}

etable_full <- dplyr::bind_rows(
    rate_row("Total (crude rate)", "total_crude"),
    rate_row("Behavioral risk factors", "behavioral"),
    rate_row("Speed-only", "speeding_only", INDENT2),
    rate_row("Alcohol-only", "alcohol_only", INDENT2),
    rate_row("Unrestrained-only", "unrestrained_only", INDENT2),
    rate_row("Speed and Alcohol", "speed_alc", INDENT2),
    rate_row("Speed and Unrestrained", "speed_unr", INDENT2),
    rate_row("Alcohol and Unrestrained", "alc_unr", INDENT2),
    rate_row("All three flagged", "all_three", INDENT2),
    rate_row("No behavioral risk factor flagged", "none_flagged"),
    rate_row("Other road users", "other"),
    rate_row("Pedestrian", "pedestrian", INDENT2),
    rate_row("Bicyclist", "bicyclist", INDENT2),
    rate_row("Other (truck/bus/ATV/etc.)", "other_nonpv", INDENT2)
)
```

## Table Notes

- Category-level US motor vehicle deaths and per-capita mortality rates, 2021 and 2024.
- Death counts are the observed FARS totals; rates are deaths per 100,000 population for each of the `r length(return_all_bins())` mutually exclusive categories, grouped into behavioral risk factors, fatalities with no behavioral flag, and other road users.
- National population denominators are shown in the year headers.
- The rate change column is the simple 2024-minus-2021 difference in per-capita rates, **not** the Kitagawa per-mile risk contribution shown in Table 1.
- FARS death counts and US Census Bureau population estimates were used to calculate crude mortality rates (2018-2024).
- 95% uncertainty intervals were calculated using the percentile method from a `r format(n_iter, big.mark = ",")`-iteration negative-binomial Monte Carlo simulation.

```{r}
band_idx <- which(
    etable_full$Category %in% c(
        "Total (crude rate)", "Behavioral risk factors",
        "No behavioral risk factor flagged", "Other road users"
    )
)

gt::gt(etable_full) |>
    gt::tab_spanner(
        label = sprintf("2021 (N=%s)", fmt_int(pop_2021)),
        columns = c(d2021, r2021)
    ) |>
    gt::tab_spanner(
        label = sprintf("2024 (N=%s)", fmt_int(pop_2024)),
        columns = c(d2024, r2024)
    ) |>
    gt::cols_label(
        d2021 = "Deaths, N", r2021 = "Rate (95% UI)",
        d2024 = "Deaths, N", r2024 = "Rate (95% UI)",
        change = "Rate change (95% UI)"
    ) |>
    gt::tab_style(
        style = list(
            gt::cell_text(weight = "bold"),
            gt::cell_fill(color = "#f0f0f0")
        ),
        locations = gt::cells_body(rows = band_idx)
    ) |>
    gt::cols_align(align = "right", columns = -1) |>
    gt::cols_align(align = "left", columns = 1) |>
    gt::tab_options(table.width = gt::pct(100))
```

```{r write-csv}
print_df <- etable_full |>
    dplyr::rename(
        `Deaths 2021, N` = d2021,
        `2021 rate (95% UI)` = r2021,
        `Deaths 2024, N` = d2024,
        `2024 rate (95% UI)` = r2024,
        `Rate change (95% UI)` = change
    )

readr::write_csv(
    print_df,
    here::here("output", "etable01_bin_rates.csv")
)
```
