---
title: "eTable 4. Weather distribution and partition of the per-mile risk effect"
format:
  html:
    embed-resources: true
    page-layout: full
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(here)
library(gt)
library(arrow)
source(here::here("code", "utils.R"))

## Non-breaking spaces so gt preserves indentation in HTML (same as Table 1).
INDENT2 <- strrep("\u00a0", 4L)

cat_label <- c(
    good = "Good (clear or cloudy)",
    adverse = "Adverse (rain/snow/sleet/hail/fog/smoke/etc.)",
    unknown = "Unknown / not reported"
)
```

```{r distribution-data}
fp <- arrow::read_parquet(
    here::here("data", "fars_persons_2018_2024.parquet")
) |>
    dplyr::filter(year %in% c(2021, 2024))

pop <- arrow::read_parquet(here::here("data", "panel_state_year.parquet")) |>
    dplyr::filter(year %in% c(2021, 2024)) |>
    dplyr::group_by(year) |>
    dplyr::summarise(pop = sum(population), .groups = "drop")
pop_2021 <- pop$pop[pop$year == 2021]
pop_2024 <- pop$pop[pop$year == 2024]

counts <- fp |> dplyr::count(year, weather_cat, name = "deaths")
tot <- function(yr) {
    sum(counts$deaths[counts$year == yr])
}
get_d <- function(wc, yr) {
    v <- counts$deaths[counts$weather_cat == wc & counts$year == yr]
    if (length(v) == 0) 0L else v
}

dist_row <- function(label, wc, indent = "") {
    d21 <- if (is.null(wc)) tot(2021) else get_d(wc, 2021)
    d24 <- if (is.null(wc)) tot(2024) else get_d(wc, 2024)
    tibble::tibble(
        Category = paste0(indent, label),
        d2021 = fmt_int(d21),
        p2021 = fmt_pct(100 * d21 / tot(2021)),
        r2021 = fmt_rate(d21, pop_2021),
        d2024 = fmt_int(d24),
        p2024 = fmt_pct(100 * d24 / tot(2024)),
        r2024 = fmt_rate(d24, pop_2024)
    )
}

dist_tbl <- dplyr::bind_rows(
    dist_row("Total (all deaths)", NULL),
    dist_row(cat_label["good"], "good", INDENT2),
    dist_row(cat_label["adverse"], "adverse", INDENT2),
    dist_row(cat_label["unknown"], "unknown", INDENT2)
)

## Adverse-of-known share by year across the analysis window (for Table Notes).
adv_share_yr <- arrow::read_parquet(
    here::here("data", "fars_persons_2018_2024.parquet")
) |>
    dplyr::filter(
        year %in% config::get("year_i"):config::get("year_j"),
        weather_cat %in% c("good", "adverse")
    ) |>
    dplyr::count(year, weather_cat) |>
    tidyr::pivot_wider(names_from = weather_cat, values_from = n) |>
    dplyr::mutate(adv_pct = 100 * adverse / (adverse + good))
adv_min <- min(adv_share_yr$adv_pct)
adv_max <- max(adv_share_yr$adv_pct)
```

## Table Notes — distribution

- Distribution of US motor vehicle deaths by crash weather, 2021 and 2024.
- Death counts are the observed FARS totals; the "% of deaths" column is each weather category's share of all motor vehicle deaths that year (good + adverse + unknown = 100%).
- Crude rate is deaths per 100,000 total US population (the denominator shown in the year headers); it is not a weather-specific population rate, because exposure (driving) is not observed by weather.
- Weather is taken from FARS `WEATHERNAME`, grouped as good (clear or cloudy), adverse (rain, snow, sleet, hail, fog, smoke, severe crosswinds, or other adverse conditions), or unknown.
- Comparing the "% of deaths" columns shows whether the weather distribution of deaths shifted between 2021 and 2024.

```{r tbl-distribution}
band_idx <- which(dist_tbl$Category == "Total (all deaths)")

gt::gt(dist_tbl) |>
    gt::tab_spanner(
        label = sprintf("2021 (N=%s)", fmt_int(pop_2021)),
        columns = c(d2021, p2021, r2021)
    ) |>
    gt::tab_spanner(
        label = sprintf("2024 (N=%s)", fmt_int(pop_2024)),
        columns = c(d2024, p2024, r2024)
    ) |>
    gt::cols_label(
        d2021 = "Deaths, N", p2021 = "% of deaths", r2021 = "Crude rate",
        d2024 = "Deaths, N", p2024 = "% of deaths", r2024 = "Crude rate"
    ) |>
    gt::tab_style(
        style = list(
            gt::cell_text(weight = "bold"),
            gt::cell_fill(color = "#f0f0f0")
        ),
        locations = gt::cells_body(rows = band_idx)
    ) |>
    gt::cols_align(align = "right", columns = -1) |>
    gt::cols_align(align = "left", columns = 1) |>
    gt::tab_options(table.width = gt::pct(100))
```

```{r decomposition-data}
pt <- arrow::read_parquet(
    here::here("data", "decomposition_weather.parquet")
) |>
    dplyr::select(weather_cat, mean = risk_contrib)

w_ci <- arrow::read_parquet(
    here::here("data", "summaries", "weather_ci.parquet")
)
n_iter <- max(w_ci$n[w_ci$geography == "national"])
ui <- w_ci |>
    dplyr::filter(geography == "national", component == "risk_contrib") |>
    dplyr::select(weather_cat, p025, p975)

risk_det <- arrow::read_parquet(
    here::here("data", "decomposition_chained.parquet")
) |>
    dplyr::filter(geography == "national", scope == "cumulative_2021_2024") |>
    dplyr::pull(risk_effect)
risk_ui <- arrow::read_parquet(
    here::here("data", "summaries", "main_ci.parquet")
) |>
    dplyr::filter(geography == "national", component == "risk_effect")

dat <- pt |>
    dplyr::left_join(ui, by = "weather_cat") |>
    dplyr::mutate(ci = fmt_ci_signed(mean, p025, p975)) |>
    dplyr::mutate(
        weather_cat = factor(weather_cat, levels = names(cat_label))
    ) |>
    dplyr::arrange(weather_cat) |>
    dplyr::transmute(
        Category = paste0(INDENT2, cat_label[as.character(weather_cat)]),
        ci
    )

total_row <- tibble::tibble(
    Category = "Per-mile risk effect (total)",
    ci = fmt_ci_signed(risk_det, risk_ui$p025, risk_ui$p975)
)

decomp_tbl <- dplyr::bind_rows(total_row, dat) |>
    dplyr::rename(`Contribution to per-mile risk effect` = ci)
```

## Table Notes — partition

- National per-mile risk effect partitioned by crash weather, 2021-2024.
- FHWA does not report VMT by weather, so this is a partition of the per-mile risk effect, **not** a stratified decomposition.
- Because the three categories exhaust all decedents, their contributions sum to the per-mile risk effect shown in the total row.
- The partition assumes a stable weather mix of driving over the window (among fatalities with known weather, the adverse share held between `r fmt_pct(adv_min)` and `r fmt_pct(adv_max)` across `r config::get("year_i")`-`r config::get("year_j")`).
- Death draws were taken negative-binomial at the state by age group by category by weather cell; method-of-moments dispersion from the 2021-2024 window (`r format(n_iter, big.mark = ",")` iterations).
- All values per 100,000 population (95% uncertainty intervals).

```{r tbl-weather}
band_idx_d <- which(decomp_tbl$Category == "Per-mile risk effect (total)")

gt::gt(decomp_tbl) |>
    gt::tab_style(
        style = list(
            gt::cell_text(weight = "bold"),
            gt::cell_fill(color = "#f0f0f0")
        ),
        locations = gt::cells_body(rows = band_idx_d)
    ) |>
    gt::cols_align(align = "right", columns = -1) |>
    gt::cols_align(align = "left", columns = 1) |>
    gt::tab_options(table.width = gt::pct(100))
```

```{r write-csv}
readr::write_csv(
    dist_tbl,
    here::here("output", "etable04_weather_distribution.csv")
)
readr::write_csv(
    decomp_tbl,
    here::here("output", "etable04_weather.csv")
)
```
